### 2020-11-13 doc/manuscript/Dow_manuscript.Rmd
cleanup, organization

## (1. forests are critical for climate change regulation, so we need to understand better how they're responding to climate change)
As global atmospheric greenhouse gas levels are rising, and little meaningful policy is being implemented in response, our planet is expected to surpass the 2.5 degree C warming mark that the Paris accord pledged to avoid. The absence of human intervention increases our reliance on natural systems to soften the blow of climate change. Plant life, especially forests, make up a large portion of these natural systems, offering many valuable ecosystem services. For one, they house an astonishing 80% of terrestrial species and are capable of removing the greenhouse gas CO2 from the atmosphere and storing it in the trees or soil. Forests sequester about one-third of all anthropogenic CO~2~ and store ~45% of all terrestrial carbon [@bonan_forests_2008]. However, their future under the current era of rapid change is uncertain [@friedlingstein_climatecarbon_2006].

@@

## (2. spring temperatures are rising, and this is affecting leaf phenology and carbon cycling)
In recent decades, an increase in average spring temperature has been impacting many natural systems including forests (Parmesan & Yohe, 2003; Zhihong, 2006). 
Shifts in leaf phenology timing could be accompanied by changes in stem-growth phenology, which would likely affect woody productivity, carbon and nutrient cycling, and water use patterns in temperate forests (Keeling et al, 1996;). 
The effects of this spring warming are most apparent when examining leaf phenology shifts, *which have been extensively studied (Way, 2011; Klosterman & Hufkins & Richardson, 2018), leading to many critical observations*. As temperatures are increasing, some regions have seen earlier leaf out in deciduous trees, sometimes matched or exceeded by later leaf senescence, meaning a potential shift in growing season timing or length (Ibáñez et al, 2010; Jeong et al., 2011). The timing of this spring warming also has a significant effect on leaf phenology response (Crabbe et al, 2016).

@@

## (3. However, we don't know how warming springs affect stem-growth, which matters because the C allocated to stem growth will be retained in the ecosystem for decades to millenia )

@@

## (4. We also know little about differences across species or functional groups, which matters because it will affect how different species fare under climate change )
~~A recent publication by D'Orangeville et al () found that~~ Leaf and wood phenology  traits, along with their apparent relationships, are dissimilar among trees of different wood types (ring-porous and diffuse-porous; **D'Orangeville et al. YY**). For example, ~~they observed that~~ bud burst occurs around the same time as 25% total annual growth, and 75% leaf development coincided with 50% total annual growth in ring-porous trees, while leaf phenology and stem growth phenology were decoupled in diffuse-porous trees.

@@

## (5. Here, we...)
Here, we characterize how early spring temperatures affect stem growth phenology and rates of temperate deciduous trees within two forest dynamics plots in the Eastern USA. Using data from dendrometer bands measured throughout the growing season, we fit a growth model  to the time series of individual trees to determine the day of year (DOY) where 25, 50, and 75% annual growth was achieved; maximum growth rates and the DOY on which they occurred; total annual growth; and peak growing season length (75%-25% DOY) [Fig. `r schematic`; @mcmahon_general_2015]. We test the hypotheses that (1) warmer early springs result in earlier stem growth and a period of growth, but (2) maximum growth rates are independent of spring temperatures, and as a result (3) total annual growth increases in response to warmer spring temperatures (Table `r hypothesis_table`).

@@

@@

(*tree rings, if included-- KAT can write this*)

@@

Daily maximum temperatures ($T_{MAX}$) for SCBI were obtained from a meteorological tower adjacent to the SCBI ForestGEO plot (data can be found on the ForestGEO climate portal, DOI on github). Daily maximum temperatures at Harvard Forest were obtained from the Harvard Forest weather station (Boose & Gould, 2004; Boose, 2018).  The R package climpact () was used to plot $T_{MAX}$ for visual inspection. Climpact also identifies readings that were greater than 3 standard deviations away from yearly means. These points were labeled as outliers and removed from the dataset. Gaps in the SCBI meteorological tower data were subsequently filled using $T_{MAX}$ readings obtained from a National Center for Environmental Information (NCEI) weather station located in Front Royal, Virginia (https://www.ncdc.noaa.gov/cdo-web/datasets/GHCND/stations/GHCND:USC00443229/detail). (*how were gaps filled for Harvard? Or was the record good?*)

## Fitting growth curves

@@

## Determining optimal climate windows

@@

## Hypothesis testing

@@

*Phenological patterns*

*(Numbers in bold will change. If we bring in HF and/or SERC, we should create a table, rather than trying to list everything in the text.)*

Across all trees and years, 25% of annual growth was achieved on **May 9** ($DOY_{25}$=**129**), 50% on **June 6** ($DOY_{50}$=**157**), and 75% on **July 3** ($DOY_{75}$=**184**), for an average peak growing season length of **56** days. Maximum growth rate occurred *just one day after $DOY_{50}$* on **June 7** ($DOY_{max}$=**158**).

Growth pheonology differed between ring- and diffuse- porous species, with ring-porous species having an earlier start to growth, with growth spread out over a longer growing season (Fig. `r growth_curves_all`).
For ring-porous trees, on average, 25% of annual growth was achieved on **April 23** ($DOY_{25}$=**113**), 50% on **May 27** ($DOY_{50}$=**147**), and 75% on **June 30** ($DOY_{75}$=**181**), for an average peak growing season length of **68** days. Maximum growth rate occurred *on $DOY_{50}$* and averaged .048 **mm day^-1^ (?)**. 
For diffuse-porous trees, on average, 25% of annual growth was achieved on **June 2** ($DOY_{25}$=**153**), 50% on **June 20** ($DOY_{50}$=**171**), and 75% on **July 9** ($DOY_{75}$=**190**), for an average peak growing season length of **37** days. Maximum growth rate occurred *just one day after $DOY_{50}$* on **June 21** ($DOY_{max}$=**172**) and averaged .068 **mm day^-1^ (?)**.

@@

*Optimal climate windows*

@@

*Climate sensitivity of phenology*

Warmer spring temperatures resulted in an earlier start of growth for ring-porous species and an earlier growing season for diffuse-porous species (Fig. `r pheno_Tsensitivity`).

@@

*Climate sensitivity of growth*

@@

>>>>

## (forests are critical for climate change regulation, so we need to understand better how they're responding to climate change)
As global atmospheric greenhouse gas levels are rising, and little meaningful policy is being implemented in response, our planet is expected to surpass the 2.5 degree C warming mark that the Paris accord pledged to avoid. The absence of human intervention increases our reliance on natural systems to soften the blow of climate change. Plant life, especially forests, make up a large portion of these natural systems, offering many valuable ecosystem services. For one, they house an astonishing 80% of terrestrial species and are capable of removing the greenhouse gas CO2 from the atmosphere and storing it in the trees or soil. Forests sequester about one-third of all anthropogenic CO~2~ and store ~45% of all terrestrial carbon [@bonan_forests_2008]. However, their future under the current era of rapid change is uncertain [@friedlingstein_climatecarbon_2006].

@@

## (spring temperatures are rising, and this is affecting leaf phenology and carbon cycling)
In recent decades, an increase in average spring temperature has been impacting many natural systems including forests (Parmesan & Yohe, 2003; Zhihong, 2006). 
Shifts in leaf phenology timing could be accompanied by changes in stem-growth phenology, which would likely affect woody productivity, carbon and nutrient cycling, and water use patterns in temperate forests (Keeling et al, 1996;). 
The effects of this spring warming are most apparent when examining leaf phenology shifts, *which have been extensively studied (Way, 2011; Klosterman & Hufkins & Richardson, 2018), leading to many critical observations*. As temperatures are increasing, some regions have seen earlier leaf out in deciduous trees, sometimes matched or exceeded by later leaf senescence, meaning a potential shift in growing season timing or length (Ibáñez et al, 2010; Jeong et al., 2011). The timing of this spring warming also has a significant effect on leaf phenology response (Crabbe et al, 2016).

@@

## (However, we don't know how warming springs affect stem-growth, which matters because the C allocated to stem growth will be retained in the ecosystem for decades to millenia )

@@

## (We also know little about differences across species or functional groups, which matters because it will affect how different species fare under climate change )
~~A recent publication by D'Orangeville et al () found that~~ Leaf and wood phenology  traits, along with their apparent relationships, are dissimilar among trees of different wood types (ring-porous and diffuse-porous; **D'Orangeville et al. **). For example, ~~they observed that~~ bud burst occurs around the same time as 25% total annual growth, and 75% leaf development coincided with 50% total annual growth in ring-porous trees, while leaf phenology and stem growth phenology were decoupled in diffuse-porous trees.

@@

## (Here, we...)
Here, we characterize how early spring temperatures affect stem growth phenology and rates of temperate deciduous trees within two forest dynamics plots in the Eastern USA. Using data from dendrometer bands measured throughout the growing season, we fit a growth model  to the time series of individual trees to determine the day of year (DOY) where 25, 50, and 75% annual growth was achieved; maximum growth rates and the DOY on which they occurred; total annual growth; and peak growing season length (75%-25% DOY) [Fig. `r schematic`; @mcmahon_general_2015]. We test the hypotheses that (1) warmer early springs result in earlier stem growth and a period of growth, but (2) maximum growth rates are independent of spring temperatures, and as a result (3) total annual growth increases in response to warmer spring temperatures (Table `r hypothesis_table`).

@@

## Data collection and cleaning
### Dendrometer bands

@@

@@

### Tree-rings
(*tree rings, if included-- KAT can write this*)

### Climate
Daily maximum temperatures ($T_{MAX}$) for SCBI were obtained from a meteorological tower adjacent to the SCBI ForestGEO plot (data can be found on the ForestGEO climate portal, DOI on github). Daily maximum temperatures at Harvard Forest were obtained from the Harvard Forest weather station (Boose & Gould, 2004; Boose, 2018).  The R package climpact () was used to plot $T_{MAX}$ for visual inspection. Climpact also identifies readings that were greater than 3 standard deviations away from yearly means. These points were labeled as outliers and removed from the dataset. Gaps in the SCBI meteorological tower data were subsequently filled using $T_{MAX}$ readings obtained from a National Center for Environmental Information (NCEI) weather station located in Front Royal, Virginia (https://www.ncdc.noaa.gov/cdo-web/datasets/GHCND/stations/GHCND:USC00443229/detail). (*how were gaps filled for Harvard? Or was the record good?*)

## Data analysis
### Fitting growth curves

@@

### Determining optimal climate windows

@@

### Hypothesis testing

@@

## Phenological patterns

*(Numbers in bold will change)*

Across all trees and years at SCBI, 
25% of annual growth was achieved on **May 9** ($DOY_{25}$=**129**), 
50% on **June 6** ($DOY_{50}$=**157**), 
and 75% on **July 3** ($DOY_{75}$=**184**), 
for an average peak growing season length of **56** days. 
Maximum growth rate occurred *just one day after $DOY_{50}$* on **June 7** ($DOY_{max}$=**158**). 
At Harvard Forest,
25% of annual growth was achieved on **XX** ($DOY_{25}$=**XX**), 
50% on **XX** ($DOY_{50}$=**XX**), 
and 75% on **XX** ($DOY_{75}$=**XX**), 
for an average peak growing season length of **XX** days. 
Maximum growth rate occurred on **XX** ($DOY_{max}$=**XX**).

At both sites, growth phenology differed between ring- and diffuse- porous species, with ring-porous species having an earlier start to growth, with growth spread out over a longer growing season (Fig. `r growth_curves_all`, Table `r mean_growth_parameters_table`).

~~For ring-porous trees, on average, 25% of annual growth was achieved on **April 23** ($DOY_{25}$=**113**), 50% on **May 27** ($DOY_{50}$=**147**), and 75% on **June 30** ($DOY_{75}$=**181**), for an average peak growing season length of **68** days. Maximum growth rate occurred *on $DOY_{50}$* and averaged .048 **mm day^-1^ (?)**.~~
~~For diffuse-porous trees, on average, 25% of annual growth was achieved on **June 2** ($DOY_{25}$=**153**), 50% on **June 20** ($DOY_{50}$=**171**), and 75% on **July 9** ($DOY_{75}$=**190**), for an average peak growing season length of **37** days. Maximum growth rate occurred *just one day after $DOY_{50}$* on **June 21** ($DOY_{max}$=**172**) and averaged .068 **mm day^-1^ (?)**.~~

@@

## Optimal climate windows

@@

## Climate sensitivity of phenology

For both ring- and diffuse- porous species at both sites, warmer spring temperatures resulted in earlier achievement of growth milestones (Tables `r hypothesis_table`, Figs. `r pheno_Tsensitivity`, `r DOYtiming`).
Specifically, $DOY_{25}$, $DOY_{50}$, $DOY_{75}$, and $DOY_{g.max}$ all decreased significantly with increasing $T_{max}$ during the critical period (Table `r mean_growth_parameters_table`).

@@

Because both $DOY_{25}$ and $DOY_{75}$ both shifted forward, peak growing season length was not consistently increased by warmer early spring temperatures (Tables `r hypothesis_table`, Figs. `r pheno_Tsensitivity`, `r DOYtiming`).
At both sites, $DOY_{75}-DOY_{25}$ decreased with temperature for ring-porous species 
and was neutral (SCBI) or positive (Harvard Forest) for diffuse-porous species.

## Climate sensitivity of growth

Maximum growth rate, $g_{max}$, was not consistently linked to early spring temperatures (Tables `r hypothesis_table`, Fig. `r pheno_Tsensitivity`). 
Ring-porous species exhibited no change (SCBI) or a slight increase (Harvard Forest) with warmer early spring temperatures.
In contrast, diffuse-porous species exhibited lower $g_{max}$ in years with higher early spring $T_{max}$.

Total annual growth rate, $\Delta DBH$, also was not consistently linked to early spring temperatures (Tables `r hypothesis_table`, Fig. `r pheno_Tsensitivity`).

@@

## (summary of key findings, gap filled)
We found that warmer spring temperatures in temperate deciduous forests cause earlier tree growth but have little effect on annual woody productivity...

## ("limitation in filling gap, how to interpret/fix")

## ("limitation in generalization, how to interpret/fix")

## ("contributions beyond")

## ("science is better now")
